Topological Equivariant Artist Model for Visualization Library Architecture

### 2021-11-11 paper/main.tex
reworked defining artist section

\begin{definition}
The \textcolor{base}{base space} \dbasec\ is a topological space, which means it is a set $\dbase$ with a collection of open sets \cite{TopologyVsTopology} surrounding each element in the set. Open sets are a collection of subsets $\{\opensetc\}$ in a set \dbase, including the empty set, the whole set \dbase, and every union and intersection of its member subsets. For each point $\dbasepoint in \dbase$, there is a collection of subsets $\\{\openset\} \subset \dbase$ such that $\dbasepoint \in \openset \subseteq \dbase$ \cite{spanier1989algebraic, bradleyTopologyCategoricalApproach2020}  
\end{definition}
\note{intuition about base space}

@@

@@

 \begin{definition} A functor is a morphism between categories $F: \mathcal{C} \rightarrow \mathcal{D}$. A functor has the properties of identity and composition \cite{riehlCategoryTheoryContext}\end{definition}
A functor preserves the structure of the category, namely morphisms and the source and target objects of those morphisms, composition of morphisms, and identity morphisms \cite{riehlCategoryTheoryContext}. Contravariant functors are functors where the morphisms in $\mathcal{C}$ go in the opposite direction from the morphisms in $\mathcal{D}$.

@@

\section{Artist} 
In this section, we use category theory to formally express the implicit assumptions that visualization library components make in transforming data into graphical representations. We propose that the visualization transformation can be modeled as a functor, which we call the \textcolor{artist}{Artist} \vartist \footnote{We call this transformation the artist because the \mintinline{python}{Artist} object in Matplotlib\cite{hunterMatplotlib2DGraphics2007}}. The artist \vartist\ converts data, which we model as the sheaf $\sheaf_{\dtotal}$ to graphics in the sheaf $\sheaf_{\gtotal}$. We formulate this association as

\begin{equation}
  \vartistc: \sheafc_{\dtotalc} \rightarrow \sheafc_{\vartistc}
  \label{eq:artist:artist}
\end{equation}

@@

The category $\mathcal{H}$ has as objects the sets of functions that generate every arbitrary graphic over the open sets in $\mathcal{\gtotal}$. While $\mathcal{H}$ is the domain of the artist \vartist, the range of \vartist\ is the subset of graphics $\sheaf_{\vartist} \subset \sheaf_{\mathcal{\dtotal}}$ such that structure imposed by $\dfunc_{\dtotal}$ commutes 
\begin{equation}
  \label{eq:artist:equivariance}
  \begin{tikzcd}
    \sheafc_{\dtotalc} \arrow[r, "\vartistc"] \arrow[d, "\dfuncc_{\dtotalc}"'] & \sheafc_{\vartistc} \arrow[d, "\dfuncc_{\vartistc}"] \\
    \sheafc_{\dtotalc} \arrow[r, "\vartistc"] & \sheafc_{\vartistc}                                              
  \end{tikzcd}
\end{equation}
The equivariance condition expressed in \label{eq:artist:equivariance} is that a graphic generated by  transforming and then visualizing data $\vartist(\dfunc_{\dtotal}(\sheaf_{\mathcal{dtotal}}))$ can equivalently be generated by visualizing the data and then transforming the graphic $\dfunc_{\vartist}(\vartist(\sheaf_{\mathcal{\dtotal}}))$. This expectation of equivariance is illustrated in \autoref{fig:related-work:equivariance}

@@

We can compose these diagrams together to generate an \vartist\ to handle any of the things.

@@

\begin{definition} Following from \label{def:artist:} the artist \vartist is a special case of  natural transformations

  %% this equation is actually on sheafs and is not a map 
  \begin{equation}
    \vartist: \Gamma(\dtotal, \dbase)\nrightarrow \Gamma(\gtotal, \gbase)
  \end{equation}
  between the profunctor objects $\Gamma(E),\Gamma(H)$ that preserves continuity and equivariance.
\end{definition}
  Natural transformations are functions that map functions between categories (functors) to other functors in a structure preserving way \cite{riehlCategoryTheoryContext, spanier1989algebraic, fongInvitationAppliedCategory2019}.

\begin{definition} We define the natural transformation \textcolor{artist}{Artist}  \vartist\ as the tuple $(\vindex, \vchannel, \vmark, \mathcal{\dtotal},\mathcal{\vtotal}, \mathcal{\gtotal})$ where
  \begin{enumerate}
    \item \vindex\ is a continuity preserving functor
    \item \vchannel\ and \vmark\ are equivariance preserving functors
    \item $\mathcal{\dtotal}$, $\mathcal{\vtotal}$, and $\mathcal{\gtotal}$ are profunctor categories of sheafs
    \item 
  \end{enumerate}
\end{definition}

%%% no arrow V->H
%%% Q is defined at the level of Gamma\begin{equation}
\begin{equation}
  \label{eq:artist:construction:all}
  \begin{tikzcd}
    \dtotalc \arrow[rr, "\vchannel"] \arrow[rdd, "\pi"] &         & \dtotalc \arrow[ldd, "\pi"'] & \mathcal{\vtotalc}                                                        & Im_{\vmarkc} \subset \gtotalc                     \\
    &         &                              & {} \arrow[r, "\vmarkc"]                                                   & {}                                                \\
    & \dbasec &                              & \mathcal{\dbasec} \arrow[uu, "\sheaf_{\vtotalc}"] \arrow[r, "\vindexc^*"] & \mathcal{\gbasec} \arrow[uu, "\sheaf_{\vmarkc}"']
    \end{tikzcd}
\end{equation}

\subsection{Data}
We model data as sections of a fiber bundle $(\dtotal, \pi, \dbase, \dfiber)$. We encode the continuity of the data as the \textcolor{base}{base space} $\dbase$. \note{does this go in intro?} The \textcolor{fiber}{fiber space} \dfiber\ is the space of all possible values of the data \autoref{eq:related-work:fiber}. We model the data as the section \dsection because, as described in \autoref{sec:related-work:fiber-bundles}, the section is the map from the indexing space \dbase to the space of possible data values \dfiber. Spivak's notation, as discussed in \autoref{sec:related-work:equivariance}, also allows for associating elements in a section with the field it comes from. This allows for field based selection of values, while inclusion allows for continuity (index) based selections.

@@

\subsection{Graphic} 
The object of the graphic category $\mathcal{\gtotal}$ is the fiber bundle \gtotal. The bundle \gtotal\ has the same structure as the data bundle \dtotal

>>>>

\begin{definition}
The \textcolor{base}{base space} \dbasec\ is a topological space, which means it is a set $\dbase$ with a collection of open sets \cite{TopologyVsTopology} surrounding each element in the set. Open sets are a collection of subsets $\{\opensetc\}$ in a set \dbase, including the empty set, the whole set \dbase, and every union and intersection of its member subsets. For each point $\dbasepoint \in \dbase$, there is a collection of subsets $\\{\openset\} \subset \dbase$ such that $\dbasepoint \in \openset \subseteq \dbase$ \cite{spanier1989algebraic, bradleyTopologyCategoricalApproach2020}  
\end{definition}
\note{intuition about base space}

@@

\note{the notation here may need to be X and Y}

@@

 \begin{definition} A functor is a morphism between categories $\mathfrak{F}: \mathcal{C} \rightarrow \mathcal{D}$. A functor has the properties of identity and composition \cite{riehlCategoryTheoryContext}\end{definition}
A functor preserves the structure of the category, namely morphisms and the source and target objects of those morphisms, composition of morphisms, and identity morphisms \cite{riehlCategoryTheoryContext}. Contravariant functors are functors where the morphisms in $\mathcal{C}$ go in the opposite direction from the morphisms in $\mathcal{D}$.

@@

\section{Artist} 
In this section, we use category theory to formally express the implicit assumptions that visualization library components make in transforming data into graphical representations. We propose that the visualization transformation can be modeled as a functor, which we call the \textcolor{artist}{Artist} \vartist \footnote{We call this transformation the artist because the \mintinline{python}{Artist} object in Matplotlib\cite{hunterMatplotlib2DGraphics2007}}. The artist \vartist\ converts data, which we model as the sheaf $\sheafc_{\dtotalc}$ to graphics in the sheaf $\sheafc_{\gtotalc}$. We formulate this association as

\begin{equation}
  \label{eq:artist:artist}
  \vartistc: \sheafc_{\dtotalc} \rightarrow \sheafc_{\gtotal}
\end{equation}

@@

The category $\mathcal{H}$ has as objects the sets of functions that generate every arbitrary graphic over the open sets in $\mathcal{\gtotal}$. While $\mathcal{H}$ is the domain of the artist \vartist, the range of \vartist\ is the subset of graphics $\sheaf_{\vartist} \subset \sheaf_{\mathcal{\dtotal}}$ such that structure imposed by $\dfunc_{\dtotal}$ commutes 
\begin{equation}
  \label{eq:artist:equivariance}
  \begin{tikzcd}
    \sheafc_{\dtotalc} \arrow[r, "\vartistc"] \arrow[d, "\dfuncc_{\dtotalc}"'] & \sheafc_{\vartistc} \arrow[d, "\dfuncc_{\vartistc}"] \\
    \sheafc_{\dtotalc} \arrow[r, "\vartistc"] & \sheafc_{\vartistc}                                              
  \end{tikzcd}
\end{equation}
The equivariance condition expressed in \autoref{eq:artist:equivariance} is that a graphic generated by  transforming and then visualizing data $\vartist(\dfunc_{\dtotal}(\sheaf_{\mathcal{dtotal}}))$ can equivalently be generated by visualizing the data and then transforming the graphic $\dfunc_{\vartist}(\vartist(\sheaf_{\mathcal{\dtotal}}))$. This is a formal statement of the equivariance illustrated in \autoref{fig:related-work:equivariance}

@@

\autoref{eq:artist:intersection:basespace} and \autoref{eq:artist:intersection:fiber} can be composed to specify how different aspects of the structure of a multivariate nested dimensional dataset, such as a spatio-temporal weather dataset, should be preserved in a visualization. 
\note{this is probably the power of this model and should maybe get a figure?}

@@

Following from \autoref{eq:artist:equivariance}, the artist \vartistc\ is a natural transformation
  \begin{equation}
    \vartist: \sheaf_{\mathcal{\dtotal}}
  \end{equation}
that we construct to preserve continuity and equivariance.

\begin{definition} We define the natural transformation \textcolor{artist}{Artist}  \vartist\ as the tuple $(\vindexc, \vchannelc, \vmarkc, \mathcal{\dtotalc},\mathcal{\vtotalc}, \mathcal{\gtotalc})$ where
  \begin{enumerate}
    \item $\vchannelc: \dtotalc \rightarrow \vtotalc$ is a bundle map from data values to the visual variables they are mapped to
    \item $\vmarkc: \sheafc_{\mathcal{\vtotalc} \rightarrow \sheafc_{\mathcal{\vartistc}}$ is a sheaf map that builds a graphic generating function parameterized by the visual variables 
    \item $\vindexc: \gbasec \rightarrow \dbasec$ is a surjective map from the graphic topological base to the data topological base
  \end{enumerate}
\end{definition}
and $\mathcal{\dtotalc}$, $\mathcal{\vtotalc}$, and $\mathcal{\gtotalc}$ are the categorical representations of fiber bundles that model the data, visual variable, and graphic space of the data to visual transformation.

\note{figure out how to format notes} Usage of the bundle directly in an equation or diagram - $\dotal, \vtotal, \gtotal$ - denotes that the function can be evaluated pointwise over $\dbasepoint \in \dbase$. Use of the categorical form -- $\mathcal{\dtotalc}, \mathcal{\vtotalc}, \mathcal{\gtotalc}$ -- denotes that the function is evaluated over an openset object, meaning that the function needs knowledge of both a value over a point and some information about neighboring values.

\begin{equation}
  \label{eq:artist:construction:all}
  \begin{tikzcd}
    \dtotalc \arrow[rr, "\vchannel"] \arrow[rdd, "\pi"'] &         & \dtotalc \arrow[ldd, "\pi"] & \mathcal{\vtotalc}                                                        & Im_{\vmarkc} \subset \gtotalc                     \\
  &         &                             & {} \arrow[r, "\vmarkc"]                                                   & {}                                \\
   & \dbasec &                             & \mathcal{\dbasec} \arrow[uu, "\sheaf_{\vtotalc}"] \arrow[r, "\vindexc^*"] & \mathcal{\gbasec} \arrow[uu, "\sheaf_{\vmarkc}"']
  \end{tikzcd}
\end{equation}

\subsection{Data Domain}
We model data as sections of a fiber bundle $(\dtotal, \pi, \dbase, \dfiber)$. We encode the continuity of the data as the \textcolor{base}{base space} $\dbase$. \note{does this go in intro?} The \textcolor{fiber}{fiber space} \dfiber\ is the space of all possible values of the data \autoref{eq:related-work:fiber}. We model the data as the section \dsection because, as described in \autoref{sec:related-work:fiber-bundles}, the section is the map from the indexing space \dbase to the space of possible data values \dfiber. Spivak's notation, as discussed in \autoref{sec:related-work:equivariance}, also allows for associating elements in a section with the field it comes from. This allows for field based selection of values, while inclusion allows for continuity (index) based selections.

@@

\subsection{Graphic Codomain} 
The object of the graphic category $\mathcal{\gtotal}$ is the fiber bundle \gtotal. The bundle \gtotal\ has the same structure as the data bundle \dtotal
